\documentclass[12pt,a4paper]{article}

% ============================================================================
% PACKAGES
% ============================================================================
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{float}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{longtable}
\usepackage{array}

\geometry{margin=1in}
\onehalfspacing

\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    breakindent=0pt,
    backgroundcolor=\color{gray!10},
    keywordstyle=\color{blue},
    commentstyle=\color{gray},
    stringstyle=\color{red},
    captionpos=b,
    showspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single
}

\begin{document}

\title{\textbf{Multi-Task Deep Learning for Diabetic Retinopathy and Diabetic Macular Edema:}\\
\large Main-Branch Architecture, Training Design, and Deployment-Oriented Methodology}
\author{RamaKrishna Sastry\\
\small National Institute of Technology Karnataka (NITK)\\
\small ramakrishna.221ds026@nitk.edu.in}
\date{April 2026}

\maketitle

\newpage
\tableofcontents
\newpage

% ============================================================================
% ABSTRACT
% ============================================================================
\begin{abstract}

\noindent \textbf{Background:} Diabetic Retinopathy (DR) and Diabetic Macular Edema (DME) are major causes of avoidable vision loss. Automated multi-task prediction from fundus images can improve screening throughput and triage quality.

\noindent \textbf{Objective:} This document presents an implementation-accurate description of the current main-branch system for joint DR (5-class) and DME (3-class) prediction, with emphasis on architecture, training dynamics, checkpoint policy, and reproducible workflow.

\noindent \textbf{Method:} The model uses a ResNet50 backbone truncated at \texttt{conv4\_block6\_out}, followed by an Atrous Spatial Pyramid Pooling (ASPP) fusion block and task-specific DR/DME heads. Training is two-stage: Stage 1 with frozen backbone and Stage 2 with controlled fine-tuning plus collapse safeguards. Joint checkpoint selection is threshold-tier based with an explicit DME floor.

\noindent \textbf{Scope Note:} This main-branch report intentionally avoids claiming finalized benchmark numbers or ablation outcomes. Quantitative experimentation and comparative studies are documented as planned future work.

\noindent \textbf{Conclusion:} The current codebase establishes a robust, clinically motivated multi-task pipeline with reproducible controls and conservative selection logic. It is well-positioned for formal experimentation and external validation in subsequent work.

\noindent \textbf{Keywords:} diabetic retinopathy, diabetic macular edema, multi-task learning, ASPP, ordinal loss, QWK, medical image analysis

\end{abstract}

\newpage

% ============================================================================
% SECTION 1
% ============================================================================
\section{Introduction}

Diabetic eye disease requires early risk stratification at scale. In practical screening systems, two requirements must be met simultaneously:

\begin{enumerate}
    \item reliable DR severity grading across five ordered classes, and
    \item reliable DME risk grading across three ordered classes.
\end{enumerate}

This project addresses both tasks in a single model to reduce deployment complexity and to leverage shared retinal representations.

\subsection{Clinical Motivation}

Manual grading is accurate but resource-intensive, especially in large screening programs and low-specialist regions. A single-network, multi-task system can support:

\begin{itemize}
    \item high-throughput first-pass triage,
    \item consistent grading behavior,
    \item specialist time prioritization for uncertain or high-risk cases.
\end{itemize}

\subsection{Paper Objective}

This paper documents the \textbf{actual main-branch implementation} rather than a hypothetical or aspirational model variant. All architecture and training details are aligned with the current code in:

\begin{itemize}
    \item \texttt{model.py}
    \item \texttt{train\_enhanced.py}
    \item \texttt{main\_pipeline.py}
    \item \texttt{config.yaml}
\end{itemize}

\newpage

% ============================================================================
% SECTION 2
% ============================================================================
\section{System Overview}

\subsection{Pipeline Stages}

The end-to-end workflow is organized as:

\begin{enumerate}
    \item \textbf{Stage 0:} data preparation and split,
    \item \textbf{Stage 1:} backbone-frozen training,
    \item \textbf{Stage 1 evaluation:} comprehensive metrics with optional calibration,
    \item \textbf{Stage 2:} fine-tuning with safeguards,
    \item \textbf{Stage 2 evaluation:} same comprehensive stack,
    \item \textbf{final aggregation:} joint-tier stage selection and pipeline report generation.
\end{enumerate}

\subsection{Model IO}

\begin{itemize}
    \item Input shape: \texttt{(512, 512, 3)}
    \item DR output head: \texttt{dr\_output} (softmax, 5 classes)
    \item DME output head: \texttt{dme\_risk} (softmax, 3 classes)
\end{itemize}

\subsection{Optimization Target}

The system is optimized and monitored using QWK-driven callbacks, with additional accuracy and error metrics. Final stage selection is not a pure single-task maximum; it follows a joint DME+DR tier policy with DME floor enforcement.

\newpage

% ============================================================================
% SECTION 3
% ============================================================================
\section{Dataset and Preprocessing}

\subsection{Primary Dataset}

The main branch is configured around IDRiD-style DR and DME labels, with support for robust CSV parsing and stratified splitting in the advanced dataset loader.

\subsection{Preprocessing and Loader Behavior}

The configured workflow includes:

\begin{itemize}
    \item target resize to \texttt{512x512},
    \item training augmentation controls,
    \item class-weight computation with clipping,
    \item optional minority oversampling for DME,
    \item split metadata persistence for reproducibility.
\end{itemize}

\subsection{Reproducibility Controls}

Global seeding is applied for Python, NumPy, and TensorFlow through the pipeline entry point. Effective runtime configuration is saved to output artifacts for experiment traceability.

\newpage

% ============================================================================
% SECTION 4
% ============================================================================
\section{Architecture (Implementation-Accurate)}

\subsection{Backbone}

The backbone is \textbf{ResNet50} with \texttt{include\_top=False}, truncated at \texttt{conv4\_block6\_out}. For input \texttt{512x512}, this yields approximately \texttt{32x32x1024} feature maps.

\textbf{Important correction:} the main branch does \underline{not} use a separate dilated residual backbone module. Multi-scale expansion is handled by ASPP after the truncated ResNet feature extractor.

\subsection{ASPP Block}

The ASPP implementation contains five branches:

\begin{itemize}
    \item 1x1 convolution branch,
    \item 3x3 atrous convolution with dilation 6,
    \item 3x3 atrous convolution with dilation 12,
    \item 3x3 atrous convolution with dilation 18,
    \item global average pooling branch projected and resized back to feature-map resolution.
\end{itemize}

All branches are concatenated and projected back to configured ASPP channels (default 256).

\subsection{Task Heads}

\subsubsection{DR Head}

The DR head is a residual MLP-style classifier with:

\begin{itemize}
    \item global average pooling,
    \item layer normalization,
    \item residual projection path,
    \item two dense blocks with batch normalization and swish activation,
    \item dropout,
    \item final softmax over 5 classes.
\end{itemize}

\subsubsection{DME Head}

The DME head is configurable:

\begin{itemize}
    \item simple dense mode, or
    \item residual MLP mode (enabled in current \texttt{config.yaml}).
\end{itemize}

It outputs a softmax distribution over 3 classes via \texttt{dme\_risk}.

\subsection{Architecture Summary Table}

\begin{table}[H]
\centering
\small
\caption{Main-Branch Model Structure}
\begin{tabular}{|p{3.2cm}|p{8.4cm}|}
\hline
\textbf{Component} & \textbf{Implementation Detail} \\
\hline
Input & \texttt{(512, 512, 3)} RGB fundus image \\
\hline
Backbone & ResNet50 truncated at \texttt{conv4\_block6\_out} \\
\hline
Fusion & ASPP with dilation rates 6, 12, 18 + 1x1 + GAP branch \\
\hline
DR head & Residual MLP classifier, output \texttt{dr\_output} (5-way softmax) \\
\hline
DME head & Configurable dense/residual MLP, output \texttt{dme\_risk} (3-way softmax) \\
\hline
Model outputs & Dictionary output: \{\texttt{dr\_output}, \texttt{dme\_risk}\} \\
\hline
\end{tabular}
\end{table}

\newpage

% ============================================================================
% SECTION 5
% ============================================================================
\section{Loss Design and Optimization}

\subsection{Multi-Task Objective}

In the compiled model, loss weights are applied as:

\begin{equation}
\mathcal{L}_{total} = \lambda_{DR}\,\mathcal{L}_{DR} + 1.0\cdot\mathcal{L}_{DME}
\end{equation}

where \(\lambda_{DR}\) corresponds to \texttt{dr\_loss\_weight} in configuration.

\textbf{Important correction:} DR is the weighted term in code. DME is the reference term with weight 1.0.

\subsection{DME Loss}

DME uses \texttt{OrdinalWeightedCrossEntropy} when enabled. It combines:

\begin{itemize}
    \item cross-entropy on (optionally) smoothed labels,
    \item focal modulation (gamma configurable),
    \item class-weight scaling,
    \item ordinal penalty weighting.
\end{itemize}

The current ordinal penalty matrix for 3 classes is:

\begin{equation}
W = \begin{bmatrix}
1.00 & 1.25 & 2.00 \\
1.25 & 1.00 & 1.25 \\
2.00 & 1.25 & 1.00
\end{bmatrix}
\end{equation}

This penalizes distant ordinal errors more strongly than adjacent errors.

\subsection{DR Loss}

DR uses categorical cross-entropy or class-weighted categorical cross-entropy depending on configuration and available class counts.

\subsection{Optimizer}

Training uses Adam with gradient clipping (\texttt{clipnorm=1.0}) in the enhanced compile path.

\newpage

% ============================================================================
% SECTION 6
% ============================================================================
\section{Two-Stage Training Strategy}

\subsection{Stage 1 (Backbone Frozen)}

Stage 1 behavior:

\begin{itemize}
    \item full model is instantiated,
    \item backbone is frozen,
    \item ASPP and both heads are trainable,
    \item class-count-informed bias initialization is applied to DR and DME output layers,
    \item QWK-aware callbacks track DME and DR validation trends.
\end{itemize}

Current configuration defaults in \texttt{config.yaml} include:

\begin{itemize}
    \item learning rate: \(2.0\times10^{-5}\),
    \item epochs: 40,
    \item DR loss weight: 0.32,
    \item focal gamma: 1.5,
    \item DME label smoothing: 0.02.
\end{itemize}

\subsection{Stage 2 (Fine-Tuning with Safeguards)}

Stage 2 behavior:

\begin{itemize}
    \item starts from selected Stage 1 checkpoint,
    \item restores weights with \texttt{skip\_mismatch=True} for compatibility,
    \item unfreezes backbone,
    \item freezes selected batch-norm subsets based on configuration,
    \item applies collapse guard and revert-if-worse protections.
\end{itemize}

Current configuration defaults include:

\begin{itemize}
    \item learning rate: \(2.0\times10^{-7}\),
    \item epochs: 50,
    \item DR loss weight: 0.13,
    \item \texttt{stage2\_freeze\_aspp\_bn=true},
    \item \texttt{stage2\_freeze\_head\_bn=true},
    \item collapse guard start epoch: 10.
\end{itemize}

\subsection{Checkpoint Types}

Per stage, the training flow can save:

\begin{itemize}
    \item \texttt{best\_qwk.weights.h5} (DME-centric),
    \item \texttt{best\_dr.weights.h5} (DR-centric),
    \item \texttt{best\_joint.weights.h5} (joint-tier compliant).
\end{itemize}

\newpage

% ============================================================================
% SECTION 7
% ============================================================================
\section{Joint Selection and Evaluation Policy}

\subsection{Joint-Tier Selection}

The main branch uses strict-to-relaxed DME/DR threshold tiers with DME floor enforcement. Primary target tier is:

\begin{equation}
\text{DME QWK} \ge 0.70 \quad \text{and} \quad \text{DR QWK} \ge 0.80
\end{equation}

If no candidate satisfies upper tiers, progressively relaxed thresholds are considered. Among candidates in the same tier, harmonic-style joint scoring is used for ranking.

\subsection{Stage Aggregation}

After stage evaluations, the pipeline builds \texttt{pipeline\_report.json} with:

\begin{itemize}
    \item per-stage raw and calibrated QWK fields,
    \item selected best stage,
    \item selection mode,
    \item best joint stage and tier metadata.
\end{itemize}

\subsection{Evaluation Stack}

Comprehensive evaluation is implemented with:

\begin{itemize}
    \item QWK, MAE, RMSE, accuracy, F1 summaries,
    \item confusion matrices,
    \item optional DME threshold calibration,
    \item optional DR threshold calibration (with accuracy-drop guard).
\end{itemize}

\newpage

% ============================================================================
% SECTION 8
% ============================================================================
\section{Main-Branch Status and Reporting Boundaries}

\subsection{What Is Included in This Branch}

This branch contains:

\begin{itemize}
    \item full model and two-stage training implementation,
    \item reproducible pipeline orchestration,
    \item structured evaluation outputs,
    \item documented checkpoint-selection logic.
\end{itemize}

\subsection{What Is Intentionally Not Claimed Here}

This manuscript does \textbf{not} present finalized, publication-grade experimental claims for:

\begin{itemize}
    \item cross-dataset benchmark superiority,
    \item ablation gains in percentage terms,
    \item fixed latency or model-size leadership claims.
\end{itemize}

Such quantitative claims require a locked experimental protocol and are deferred to future work.

\subsection{Reproducible Artifact Set}

A standard pipeline run emits artifacts including:

\begin{itemize}
    \item \texttt{effective\_config.json},
    \item stage histories and QWK logs,
    \item stage checkpoints,
    \item evaluation JSON files,
    \item final \texttt{pipeline\_report.json}.
\end{itemize}

\subsection{Best Available Logged Results (Notebook Run)}

The attached Kaggle notebook run logs provide a stronger result snapshot than the
local placeholder artifacts in this workspace. The values below are copied from
explicit log lines only.

\begin{table}[H]
\centering
\small
\caption{Verified Metrics from Notebook Logs}
\begin{tabular}{|p{3.0cm}|p{2.7cm}|p{2.2cm}|p{2.2cm}|}
\hline
\textbf{Stage / Report Point} & \textbf{Selection Context} & \textbf{DME QWK} & \textbf{DR QWK} \\
\hline
Stage 1 (raw) & Selected \texttt{best\_joint} checkpoint & 0.7089 & 0.7571 \\
\hline
Stage 1 (calibrated) & DME calibration applied; DR calibration rejected by accuracy guard & 0.7981 & 0.7571 \\
\hline
Stage 2 (raw) & Selected \texttt{best\_joint} checkpoint & 0.7398 & 0.7646 \\
\hline
Stage 2 (calibrated) & Both DME and DR calibrations applied & 0.8112 & 0.8311 \\
\hline
Pipeline summary (raw) & Best joint stage reported by pipeline & 0.7398 & 0.7646 \\
\hline
\end{tabular}
\end{table}

Additional log-derived notes:
\begin{itemize}
    \item Stage 1 peak per-epoch raw DME QWK reached 0.7926 (epoch 31), while Stage 1 peak DR QWK reached 0.7571 (epoch 40).
    \item Stage 2 peak per-epoch raw DME QWK reached 0.7398 (epoch 2), while Stage 2 peak DR QWK reached 0.7893 (epoch 27).
    \item Pipeline summary reported \texttt{best\_stage=stage2}, \texttt{selection\_mode=joint\_tiers}, and elapsed time \texttt{3456.5} seconds.
\end{itemize}

\textbf{Interpretation:} In this run, raw metrics did not cross 0.80 simultaneously for both tasks under the selected joint checkpoint, but calibrated evaluation for Stage 2 crossed 0.80 on both tasks.

\subsection{Backbone Initialization Progression in Project Runs}

The project followed a practical backbone-initialization progression:

\begin{enumerate}
    \item ResNet50 with default ImageNet initialization (raw baseline mode),
    \item EyePACS-only backbone transfer,
    \item all-datasets backbone transfer (used through \texttt{--use-eyepacs} with an all-datasets pre-trained backbone file).
\end{enumerate}

\begin{table}[H]
\centering
\small
\caption{Initialization Progression and Advantages}
\begin{tabular}{|p{3.0cm}|p{4.1cm}|p{6.2cm}|}
\hline
\textbf{Initialization phase} & \textbf{What changed} & \textbf{Observed / practical advantage} \\
\hline
Raw ResNet50 (ImageNet) & Generic ImageNet initialization with no retinal-domain pretraining artifact. & Provides a stable baseline and compatibility fallback; useful for sanity checks and for separating architecture effects from transfer-learning effects. \\
\hline
EyePACS-only backbone & Backbone initialized from fundus-domain pretraining (EyePACS source). & Better domain alignment to retinal texture/lesion statistics than pure ImageNet; improves starting point for DR/DME fine-tuning and usually reduces early optimization mismatch. \\
\hline
All-datasets backbone (current best logged run path) & Backbone loaded from the all-datasets pre-trained artifact via \texttt{--use-eyepacs}. & Strongest available transfer prior in project workflow; in verified notebook logs it yielded the best reported calibrated stage metrics (Stage 2: DME QWK 0.8112, DR QWK 0.8311) with stage2 selected by joint tiers. \\
\hline
\end{tabular}
\end{table}

\textbf{Reporting note:} Quantified metrics in this repository are fully available for the all-datasets initialization run. For ImageNet-only and EyePACS-only phases, this document records the progression and practical advantages while keeping hard numeric claims restricted to verified logs.

\newpage

% ============================================================================
% SECTION 9
% ============================================================================
\section{Challenges, Design Responses, and Practical Constraints}

\subsection{Class Imbalance}

DME is typically minority-heavy in real datasets. The code addresses this through class weighting, optional oversampling, and focal modulation in the DME loss.

\subsection{Ordinal Boundary Sensitivity}

Adjacent-class confusion is clinically common. The ordinal matrix and label smoothing are designed to improve boundary robustness while retaining gradient stability.

\subsection{Stage 2 Instability Risk}

Fine-tuning can degrade one task while improving the other. Collapse guards, delayed guard activation, BN freezing controls, and revert-if-worse checks reduce this risk.

\subsection{Generalization Limits}

Domain shift across cameras, geographies, and labeling conventions remains a practical limitation. The current branch provides the necessary hooks for systematic external validation but does not yet claim final external benchmark conclusions.

\newpage

% ============================================================================
% SECTION 10
% ============================================================================
\section{Future Scope and Planned Experimentation}

Per project direction, experimentation content is placed under future scope.

\subsection{Planned Quantitative Experiments}

\begin{enumerate}
    \item \textbf{Primary benchmark run set:} repeated-seed evaluation for stable confidence intervals.
    \item \textbf{Ablation studies:} ASPP on/off, residual-DME-head on/off, ordinal-loss variants, calibration on/off.
    \item \textbf{Selection-policy stress tests:} compare DME-only, DR-only, and joint-tier checkpoint policies.
    \item \textbf{Stage 2 stability study:} collapse guard and BN-freeze option sweeps.
    \item \textbf{Cross-dataset validation:} controlled protocol on external datasets with explicit mapping rules.
\end{enumerate}

\subsection{Clinical and Deployment Extensions}

\begin{enumerate}
    \item calibration quality and uncertainty estimation (confidence-aware triage),
    \item explainability overlays (for specialist audit workflows),
    \item model compression and edge deployment profiling,
    \item fairness and subgroup performance audits.
\end{enumerate}

\subsection{Expected Deliverables in the Experimental Branch}

\begin{itemize}
    \item frozen train/validation/test protocol,
    \item statistical significance analysis,
    \item final performance tables and confusion analyses,
    \item reproducible scripts for result regeneration.
\end{itemize}

\newpage

% ============================================================================
% SECTION 11
% ============================================================================
\section{Conclusion}

This paper now reflects the current main-branch reality of the project: a robustly engineered multi-task DR+DME pipeline built around a ResNet50 conv4 backbone, ASPP-based multi-scale fusion, and task-specific output heads with QWK-oriented training controls.

The most important technical corrections are:

\begin{itemize}
    \item architecture is ResNet50-truncated plus ASPP (not a separate DRN backbone claim),
    \item multi-task loss weighting is implemented as \(\lambda_{DR}\mathcal{L}_{DR} + \mathcal{L}_{DME}\),
    \item ordinal DME penalties follow the implemented matrix with stronger distant-class penalties,
    \item checkpoint and final stage selection are joint-tier based with explicit DME floor safeguards.
\end{itemize}

From a software-engineering perspective, the branch is mature in structure: it has reproducible configuration handling, stage-wise safety mechanisms, and comprehensive artifact generation. From a scientific-reporting perspective, this manuscript intentionally avoids unsupported benchmark claims and instead provides a faithful technical baseline for upcoming formal experimentation.

In summary, the project is in a strong transition state: implementation is stable and clinically motivated, while rigorous comparative evidence is clearly scoped, planned, and ready to be executed under a dedicated experimental protocol.

\end{document}
